---
title: "Final Presentation"
author: "Kevin Garcia"
date: "2024-04-25"
output: 
  html_document:
    code_folding: hide
---





```{r}
library(plotly)
library(ggplot2)
library(dplyr)
library(readr)
library(kableExtra)
library(stringr)
```



loading the different sets of data that will be used

```{r}
drugs_data = read.csv("Drug_overdose_death_rates__by_drug_type__sex__age__race__and_Hispanic_origin__United_States.csv")

death_rates <- read.csv("Death_rates_for_suicide__by_sex__race__Hispanic_origin__and_age__United_States.csv")

leading_causes_d = read.csv('NCHS_-_Leading_Causes_of_Death__United_States.csv')

suicide_data <- read_csv("suicide-rate-by-country-2024.csv")

california_data <- read.csv("suicide-lghc-indicator-21.csv")


```


# Questions:

1. What are the highest causes of death and how do they compare to suicide deaths?
2. What is the country with the highest suicide death rate?
3. Focusing on the United States what is the State with the highest suicide rate?
4. What is the city?
5. What are the different types of drugs being used?
6. What years had the highest suicide rates?
7. What age group has the highest suicide rates?
8. What gender has the highest suicide rates?





# DATA REGARDING SUICIDES

```{r}


# suicide_data <- read_csv("suicide-rate-by-country-2024.csv") 
# 
# 
# country_rate <- suicide_data %>%
#   select(country, Combined_Suicide_Rate = suicideRateByCountry_rate2019both) %>%
#   arrange(desc(Combined_Suicide_Rate)) %>%
#   mutate(Rank = row_number())  # Add a ranking column after arranging
# 
# table_output <- kable(country_rate, format = "html", col.names = c("Rank", "Country", "Suicide Rate (2019)"), 
#                       caption = "Ranked Suicide Rates by Country (2019)") %>%
#   kable_styling(full_width = FALSE, position = "center") %>%
#   scroll_box(width = "100%", height = "500px")
# 
# # Output the table
# print(table_output)


```


# The United States is number 23 in the list

- How does suicide compare against other causes of death



```{r}


suicides <- filter(leading_causes_d, Cause.Name == "Suicide")
heart_disease <- filter(leading_causes_d, Cause.Name == "Heart disease")
cancer <- filter(leading_causes_d, Cause.Name == "Cancer")



suicide_stats <- suicides %>%
  group_by(Year) %>%
  summarize(Total_Deaths = sum(Deaths))

heart_disease_stats <- heart_disease %>%
  group_by(Year) %>%
  summarize(Total_Deaths = sum(Deaths))

cancer_stats <- cancer %>%
  group_by(Year) %>%
  summarize(Total_Deaths = sum(Deaths))


ggplot() +
  geom_line(data = suicide_stats, aes(x = Year, y = Total_Deaths, color = "Suicide")) +
  geom_line(data = heart_disease_stats, aes(x = Year, y = Total_Deaths, color = "Heart Disease")) +
  geom_line(data = cancer_stats, aes(x = Year, y = Total_Deaths, color = "Cancer")) +
  labs(title = "Death Trends by Cause", x = "Year", y = "Total Deaths") + 
  theme_bw()


```


# This is what happens when you isolate the data for suicides throughout the years



```{r}


# Filter data for suicides
suicide_data <- leading_causes_d[leading_causes_d$Cause.Name == "Suicide",]

# Summing up deaths by year
suicide_trend <- aggregate(Deaths ~ Year, data = suicide_data, sum)

# Plotting the data
ggplot(suicide_trend, aes(x = Year, y = Deaths)) +
  geom_line(group=1, colour="blue") +
  geom_point(colour="red") +
  labs(title = "Suicides Over the Years in the United States",
       x = "Year",
       y = "Number of Suicides") +
  theme_minimal()


```






# Now let's look at the State with the highest suicide rate

```{r}

state_suicides <- leading_causes_d %>%
  filter(Cause.Name== "Suicide", State != "United States") %>%
  group_by(State) %>%
  summarise(Total_Suicides = sum(Deaths)) %>%
  arrange(desc(Total_Suicides))

# Create the bar chart with the filtered data
ggplot(state_suicides, aes(x = reorder(State, Total_Suicides), y = Total_Suicides)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  
  labs(title = "Total Suicides by State",
       x = "Total Suicides",
       y = "State") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 5), 
    plot.margin = unit(c(1, 1, 1, 1), "cm") 
  )


```


# California is the state with the highest suicide rate

- what about the city

```{r}
average_rates <- california_data %>%
  group_by(Geography) %>%
  summarise(Average_Rate = mean(Rate, na.rm = TRUE)) %>%
  arrange(desc(Average_Rate))

# Plot average rates by Geography
ggplot(average_rates, aes(x = reorder(Geography, -Average_Rate), y = Average_Rate)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Average Suicide Rates by City in California",
       x = "City",
       y = "Average Rate per 100,000 People") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



# DRUG RELATED DATA


```{r}

drugs_data %>% 
  filter(STUB_NAME == "Age") %>% 
  ggplot(aes(x = YEAR, y= ESTIMATE, color = STUB_LABEL )) + geom_line() + facet_wrap(~ as.factor(PANEL))


```


## Will be focusing on the two types of drugs being used that have the most movement

- First

```{r}
# Filter and plot data for "Drug overdose deaths involving other synthetic opioids (other than methadone)"
synthetic_opioids_data <- drugs_data %>%
  filter(STUB_NAME == "Age", PANEL == "Drug overdose deaths involving other synthetic opioids (other than methadone)") %>%
  ggplot(aes(x = YEAR, y = ESTIMATE, color = STUB_LABEL)) +
  geom_line() +
  labs(title = "Overdose Deaths Involving Other Synthetic Opioids (Excluding Methadone)",
       x = "Year",
       y = "Estimate",
       color = "Group") +
  theme_minimal()

print(synthetic_opioids_data)
```

-second

```{r}


any_opioid_data <- drugs_data %>%
  filter(STUB_NAME == "Age", PANEL == "Drug overdose deaths involving any opioid") %>%
  ggplot(aes(x = YEAR, y = ESTIMATE, color = STUB_LABEL)) +
  geom_line() +
  labs(title = "Overdose Deaths Involving Any Opioid",
       x = "Year",
       y = "Estimate",
       color = "Group") +
  theme_minimal()


print(any_opioid_data)


```





# The peak years


```{r}


# Load and prepare the dataset
drugs_data <- read_csv("Drug_overdose_death_rates__by_drug_type__sex__age__race__and_Hispanic_origin__United_States.csv") %>%
  rename(DrugType = PANEL) %>%
  mutate(
    SimpleDrugType = case_when(
      DrugType == "Drug overdose deaths involving other synthetic opioids (other than methadone)" ~ "Synthetic Opioids (Excl. Methadone)",
      DrugType == "Drug overdose deaths involving any opioid" ~ "Any Opioid",
      TRUE ~ DrugType
    ),
    Gender = case_when(
      str_detect(STUB_LABEL, "Male:") ~ "Male",
      str_detect(STUB_LABEL, "Female:") ~ "Female",
      TRUE ~ "Unspecified"
    )
  )

# Calculate and filter for peak years in a generic way for reuse
total_by_year <- drugs_data %>%
  filter(SimpleDrugType %in% c("Synthetic Opioids (Excl. Methadone)", "Any Opioid")) %>%
  group_by(SimpleDrugType, YEAR) %>%
  summarise(Total_Estimate = sum(ESTIMATE, na.rm = TRUE), .groups = 'drop')

peak_years <- total_by_year %>%
  group_by(SimpleDrugType) %>%
  slice_max(Total_Estimate) %>%
  ungroup()

peak_data_analysis <- drugs_data %>%
  filter(
    SimpleDrugType %in% peak_years$SimpleDrugType,
    YEAR %in% peak_years$YEAR
  )

# Prepare the common part for different racial analyses
peak_data_analysis <- peak_data_analysis %>%
  mutate(Race = case_when(
    str_detect(STUB_LABEL, "White") ~ "White",
    str_detect(STUB_LABEL, "Black or African American") ~ "Black or African American",
    str_detect(STUB_LABEL, "American Indian or Alaska Native") ~ "American Indian or Alaska Native",
    str_detect(STUB_LABEL, "Asian or Pacific Islander") ~ "Asian or Pacific Islander",
    str_detect(STUB_LABEL, "Hispanic or Latino") ~ "Hispanic or Latino",
    TRUE ~ "Other"
  ))

```


# PLOTING OF THE DIFFERENT RACES IN PEAK YEARS



```{r}

ggplot(peak_data_analysis, aes(x = Race, y = ESTIMATE, fill = Race)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~SimpleDrugType, scales = "free_x") +
  labs(title = "Drug Overdose Deaths by Race in Peak Years",
       x = "Race",
       y = "Total Estimates",
       fill = "Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

# WHITE RACE BY GENDER


```{r}
# Filter data for White race by gender
white_data_by_gender <- peak_data_analysis %>%
  filter(Race == "White") %>%
  group_by(SimpleDrugType, Gender, YEAR) %>%
  summarise(Total_Deaths = sum(ESTIMATE, na.rm = TRUE), .groups = 'drop')

# Filter data for Black or African American race by gender
black_data_by_gender <- peak_data_analysis %>%
  filter(Race == "Black or African American") %>%
  group_by(SimpleDrugType, Gender, YEAR) %>%
  summarise(Total_Deaths = sum(ESTIMATE, na.rm = TRUE), .groups = 'drop')

# Calculate total deaths across all groups and years for each drug type
total_deaths_by_drug <- peak_data_analysis %>%
  group_by(SimpleDrugType) %>%
  summarise(Total_Deaths = sum(ESTIMATE, na.rm = TRUE), .groups = 'drop')

```



```{r}
ggplot(white_data_by_gender, aes(x = YEAR, y = Total_Deaths, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~SimpleDrugType) +
  labs(title = "Total Deaths by Year for White Race by Gender",
       x = "Year",
       y = "Total Deaths",
       fill = "Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```



# BLACK RACE BY GENDER


```{r}
ggplot(black_data_by_gender, aes(x = YEAR, y = Total_Deaths, fill = Gender)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~SimpleDrugType) +
  labs(title = "Total Deaths by Year for the African American Race by Gender",
       x = "Year",
       y = "Total Deaths",
       fill = "Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


# Total deaths across both groups 



```{r}

ggplot(total_deaths_by_drug, aes(x = SimpleDrugType, y = Total_Deaths, fill = SimpleDrugType)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Deaths Across Both Groups by Drug Type",
       x = "Drug Type",
       y = "Total Deaths",
       fill = "Drug Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```





# DENSITY GRAHP OF GENDER THROUGHOUT THE YEARS

```{r}
# Load the dataset
drugs_data <- read_csv("Drug_overdose_death_rates__by_drug_type__sex__age__race__and_Hispanic_origin__United_States.csv")

# Rename 'PANEL' to 'DrugType' and create simplified names
drugs_data <- drugs_data %>%
  rename(DrugType = PANEL) %>%
  mutate(
    SimpleDrugType = case_when(
      DrugType == "Drug overdose deaths involving other synthetic opioids (other than methadone)" ~ "Synth. Opioids (excl. Methadone)",
      DrugType == "Drug overdose deaths involving any opioid" ~ "Any Opioid",
      TRUE ~ "Other Types"
    ),
    Gender = case_when(
      str_detect(STUB_LABEL, "Male: White") ~ "Male",
      str_detect(STUB_LABEL, "Female: White") ~ "Female",
      TRUE ~ "Unspecified"
    )
  )

# Filter data specifically for 'White' race and specified drug types
filtered_data <- drugs_data %>%
  filter(
    SimpleDrugType %in% c("Synth. Opioids (excl. Methadone)", "Any Opioid"),
    str_detect(STUB_LABEL, "White"),
    Gender %in% c("Male", "Female")
  )

# Plotting density of ESTIMATE over the years, combining male and female in the same graph for each drug type
ggplot(filtered_data, aes(x = ESTIMATE, fill = Gender)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~SimpleDrugType) +
  labs(title = "Combined Density of Drug Overdose Deaths Over Years by Gender, White",
       x = "Total Deaths Estimate",
       y = "Density") +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "red")) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text.x = element_text(size = 12),
    legend.position = "top"
  )

# Print the plot
print(ggplot)

```



# DATA THAT INCLUDES A WIDER RANGE OF YEARS FROM 1950 TO 2020

```{r}
age_specific_data <- death_rates %>%
  filter(STUB_NAME == "Age", AGE != "All ages")


p <- ggplot(age_specific_data, aes(x = YEAR, y = ESTIMATE, color = AGE)) +
  geom_line() +
  geom_point() +
  labs(title = "Interactive Plot of Suicide Rates by Age Group",
       x = "Year",
       y = "Suicide Rate per 100,000 Population",
       color = "Age Group")
  

ggplotly(p)
```



# The age groups that have the highest suicides rates throughout the years


```{r}

# Calculate the average suicide rate for each age group and sort them in descending order
average_age_rates <- death_rates %>%
  group_by(AGE) %>%
  summarise(AverageRate = mean(ESTIMATE, na.rm = TRUE)) %>%
  arrange(desc(AverageRate))

# Extract the top 3 age groups with the highest average suicide rates
top_groups <- head(average_age_rates, 5)

ggplot(top_groups, aes(x = AGE, y = AverageRate, fill = AGE)) +
  geom_bar(stat = "identity") + # Use pre-calculated heights for bars
  labs(title = "Top 5 Age Groups with the Highest Average Suicide Rates",
       x = "Age Group",
       y = "Average Suicide Rate") +
  theme_minimal() +
  theme(legend.position = "none") 
  


```

